Data Science Applied to Ag - Final Project - ML

### 2025-04-26 code/Draft_script_No_Cat_Features.qmd
d

trait_clean <- trait_df %>%
  select(-block) %>%
  mutate(
    site = str_remove_all(site, "[a-z]"), # removing all lowercase letters from site
    site = str_replace(site, "-.*$", ""), # keeping only text before any dash
    site = str_replace(site, "_.*$", "") # keeping only text before any underscore
  ) %>%
  group_by(year, site, hybrid) %>%
  summarize(
    yield_mg_ha    = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    date_planted   = first(date_planted),
    date_harvested = first(date_harvested),
    .groups = "drop"
  ) %>%
  mutate(
    adjusted_yield = adjust_yield(yield_mg_ha, grain_moisture),
    planting_date  = as.Date(date_planted,   "%m/%j/%y"),
    harvest_date   = as.Date(date_harvested, "%m/%j/%y"),
    plant_doy      = yday(planting_date),
    harvest_doy    = yday(harvest_date),
  ) %>%
  select(-yield_mg_ha, -grain_moisture, -date_planted, -date_harvested, -planting_date, -harvest_date) %>%
  ungroup()

@@

meta_df2 <- meta_df %>%
  rename(
    lon = longitude,
    lat = latitude
  ) %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  distinct(year, site, .keep_all = TRUE) %>%
  ungroup()

meta_df2

The code below will create categories for previous crops.

meta_clean <- meta_df2 %>%
  mutate(
    # 1) lowercase & trim
    prev_crop_raw = str_to_lower(str_squish(previous_crop)),
    # 2) recode into your desired levels
    prev_crop = case_when(
      prev_crop_raw == "cotton"                                   ~ "cotton",
      prev_crop_raw == "peanut"                                   ~ "peanut",
      prev_crop_raw %in% c("soybean", "soybeans")                 ~ "soybean",
      prev_crop_raw == "sorghum"                                  ~ "sorghum",
      prev_crop_raw %in% c("wheat", "winter wheat", "2019/20 wheat") ~ "wheat",
      TRUE                                                         ~ "others"
    ) %>%
    # 3) make it a factor with the exact ordering you want
    factor(levels = c("cotton", "peanut", "soybean", "sorghum", "wheat", "others"))
  ) %>%
  select(-previous_crop, -prev_crop_raw) %>%
  clean_names()

@@

```{r}
summary(trait_clean)
```
```{r}
summary(soil_clean)
```

@@

fe_month <- fieldweather %>%
  # Selecting needed variables
  dplyr::select(year, 
                site, 
                hybrid, 
                lon, 
                lat, 
                yday, 
                yield = adjusted_yield, 
                prev_crop,
                plant_doy,
                harvest_doy,
                soil.ph = soilpH, 
                soil.om.pct = om_pct, 
                soil.k.ppm = soilk_ppm, 
                soil.p.ppm = soilp_ppm, 
                dayl.s = dayl_s, #to rename variable name from "dayl_s" to dayl.s
                prcp.mm = prcp_mm_day, #to rename variable name to "prcp.mm"
                srad.wm2 = srad_w_m_2,#to rename variable name to "srad.wm2"
                tmax.c = tmax_deg_c, #to rename variable name to "tmax.c"
                tmin.c = tmin_deg_c,#to rename variable name to "tmin.c"
                vp.pa = vp_pa #to rename variable name to "vp.pa"
                ) %>%
  # Creating a date class variable  
  mutate(date_chr = paste0(year, "/", yday)) %>% 
  mutate(date = as.Date(date_chr, "%Y/%j")) %>% 
  # Extracting month from date  
  mutate(month = month(date)) %>% 
  mutate(month_abb = month(date, label = T)) #To get abbreviated month name e.g., Jan, Feb, Mar,...,Dec #month_abb is in "ord" (ordinal) format

@@

The following code chunk will feature engineer planting and harvest stages based on the following cut-offs:

Planting
•	early: DOY ≤ 120
•	normal: 121–150
•	late: > 150

Harvest
•	early: DOY ≤ 275
•	normal: 276–305
•	late: > 305

fe_month <- fe_month%>%
  mutate(
    planting_stages = case_when(
      plant_doy <= 120                     ~ "early",
      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
      plant_doy > 150                      ~ "late"
    ) %>% 
      fct_relevel("early","normal","late"),

    harvest_stages = case_when(
      harvest_doy <= 275                    ~ "early",
      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
      harvest_doy > 305                     ~ "late"
    ) %>% 
      fct_relevel("early","normal","late")
  ) %>%
  select(-plant_doy, -plant_doy)

fe_month

@@

fe_month_sum <- fe_month %>%
  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame 
  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
  summarise(across(.cols = c(soil.ph, 
                             soil.om.pct, 
                             soil.k.ppm, 
                             soil.p.ppm,
                             dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}"
                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
            ) %>%
  ungroup() #To convert from "group" to "tibble"

@@

fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
  mutate(varname = paste0(name, "_", month_abb)) %>% 
  dplyr::select(-name, -month_abb) %>% 
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(10:ncol(.), ~round(., 1) )) %>%
  select(-lon, -lat, -mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)

@@

View(fe_month_sum_wide)

@@

#write_csv(fe_month_sum_wide,
          # "../data/weather_monthsum.csv")

@@

```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
#library(caret)       # Other great library for Machine Learning 
```

@@

Creating a recipe is an easy way to port your processing steps for other data sets without needing to repeat code, and also only considering the data it is being applied to.

You can find all available recipe step options here: https://tidymodels.github.io/recipes/reference/index.html

Different model types require different processing steps.\
Let's check what steps are required for an elastic net model (linear_reg). We can search for that in this link: https://www.tmwr.org/pre-proc-table #Don't need for numeric data

@@

# Create recipe for data preprocessing
weather_recipe <- recipe(yield ~ ., data = weather_train) %>% # Remove identifier columns and months not in growing season
  step_rm(
    year,       # Remove year identifier
    site,       # Remove site identifier
    matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
  ) %>%
  # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
  step_dummy(all_nominal_predictors()) %>%
  # Imputation for any numeric NAs   
  step_impute_median(all_numeric_predictors())

@@

# Examine preprocessing steps
weather_prep
```
## 2. Training

@@

xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),  # Maximum depth of each tree
  min_n = tune(),  # Minimum samples required to split a node
  learn_rate = tune()) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
  set_engine("xgboost") %>% #specify engine 
  set_mode("regression")  # Set to mode
   # Total number of boosting iterations

        # Step size shrinkage for each boosting step

@@

```
### c. Hyperparameter grid with Latin Hypercube Sampling

@@

xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  size = 10
)

xgb_grid
```

@@

#parallel processing
registerDoParallel(cores = parallel::detectCores()-1)

@@

stopImplicitCluster()

@@

>>>>

trait_clean <- trait_df %>%
  select(-block) %>%
  mutate(
    site = str_remove_all(site, "[a-z]"), # removing all lowercase letters from site
    site = str_replace(site, "-.*$", ""), # keeping only text before any dash
    site = str_replace(site, "_.*$", "") # keeping only text before any underscore
  ) %>%
  group_by(year, site, hybrid) %>%
  summarize(
    yield_mg_ha    = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    adjusted_yield = adjust_yield(yield_mg_ha, grain_moisture),
  ) %>%
  select(-yield_mg_ha, -grain_moisture) %>%
  ungroup()

@@

meta_clean <- meta_df %>%
  rename(
    lon = longitude,
    lat = latitude
  ) %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  distinct(year, site, .keep_all = TRUE) %>%
  select(-previous_crop) %>%
  clean_names() %>%
  ungroup()

@@

```{r}
summary(trait_clean)
```

```{r}
summary(soil_clean)
```

@@

fe_month <- fieldweather %>%
  # Selecting needed variables
  dplyr::select(year, 
                site, 
                hybrid, 
                lon, 
                lat, 
                yday, 
                yield = adjusted_yield, 
                soil.ph = soilpH, 
                soil.om.pct = om_pct, 
                soil.k.ppm = soilk_ppm, 
                soil.p.ppm = soilp_ppm, 
                dayl.s = dayl_s, #to rename variable name from "dayl_s" to dayl.s
                prcp.mm = prcp_mm_day, #to rename variable name to "prcp.mm"
                srad.wm2 = srad_w_m_2,#to rename variable name to "srad.wm2"
                tmax.c = tmax_deg_c, #to rename variable name to "tmax.c"
                tmin.c = tmin_deg_c,#to rename variable name to "tmin.c"
                vp.pa = vp_pa #to rename variable name to "vp.pa"
                ) %>%
  # Creating a date class variable  
  mutate(date_chr = paste0(year, "/", yday)) %>% 
  mutate(date = as.Date(date_chr, "%Y/%j")) %>% 
  # Extracting month from date  
  mutate(month = month(date)) %>% 
  mutate(month_abb = month(date, label = T)) #To get abbreviated month name e.g., Jan, Feb, Mar,...,Dec #month_abb is in "ord" (ordinal) format

@@

@@

fe_month_sum <- fe_month %>%
  group_by(year, site, hybrid, month_abb, yield) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "yield" in the group_by() to include it in the data frame because "yield" is our response variable so we must keep it in the data frame 
  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
  summarise(across(.cols = c(soil.ph, 
                             soil.om.pct, 
                             soil.k.ppm, 
                             soil.p.ppm,
                             dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}"
                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
            ) %>%
  ungroup() #To convert from "group" to "tibble"

@@

fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
  mutate(varname = paste0(name, "_", month_abb)) %>% 
  dplyr::select(-name, -month_abb) %>% 
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(c(4:ncol(.)), ~round(., 1) )) %>%
  select( -mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)

@@

@@

write_csv(fe_month_sum_wide,
           "../data/weather_monthsum.csv")

@@

```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
#install.packages("caret")

library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
library(caret)       # Other great library for Machine Learning 
```

# Load the data set

```{r weather}

#weather <- read_csv("../data/weather_monthsum.csv")

#weather

@@

@@

# Create recipe for data preprocessing
weather_recipe <- recipe(yield ~ ., data = weather_train) %>% 
  # Remove identifier columns and months not in growing season
  step_rm(
    year,       # Remove year identifier
    site,       # Remove site identifier
    hybrid,     # Remove site identifier
    matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
  )

@@

# Examine preprocessing steps
weather_prep
```

## 2. Training

@@

xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),  # Maximum depth of each tree
  min_n = tune(),  # Minimum samples required to split a node
  learn_rate = tune()
  ) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
  set_engine("xgboost") %>% #specify engine 
  set_mode("regression")  # Set to mode

@@

### c. Hyperparameter grid with Latin Hypercube Sampling

@@

xgb_grid <- grid_latin_hypercube(
  trees(),
  tree_depth(),
  min_n(),
  learn_rate(),
  size = 20
)

xgb_grid

Note: the following code chunk would not work if "tree_depth" and "learn_rate" were not fine tuned in the "xgb_spec" object.

@@

#parallel processing
#registerDoParallel(cores = parallel::detectCores()-1) #starts parallel processing

@@

#stopImplicitCluster() #ends parallel processing

@@

## 4. Select Best Models

We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:

```{r}
# Based on lowest RMSE
best_rmse <- xgb_res %>% 
  select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse

```{r}
# Based on lowers RMSE within 1% loss
best_rmse_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rmse",
                     limit = 1
                     )%>% 
  mutate(source = "best_rmse_pct_loss")

best_rmse_pct_loss
```

```{r}
# Based on lowest RMSE within 1 se
best_rmse_one_std_err <- xgb_res %>% 
  select_by_one_std_err(metric = "rmse",
                        eval_time = 100,
                        trees
                        )%>% 
  mutate(source = "best_rmse_one_std_err")

best_rmse_one_std_err
```

Here we use all three methods which we learn in this class for R2.

```{r}
# Based on greatest R2
best_r2 <- xgb_res %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```

```{r}
# Based on lowers R2 within 1% loss
best_r2_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rsq",
                     limit = 1
                     ) %>% 
  mutate(source = "best_r2_pct_loss")

best_r2_pct_loss
```

```{r}
# Based on lowest R2 within 1 se
best_r2_one_std_error <- xgb_res %>% 
  select_by_one_std_err(metric = "rsq",
                        eval_time = 100,
                        trees
                        ) %>%
  mutate(source = "best_r2_one_std_error")

best_r2_one_std_error
```

## Compare and Finalize Model

```{r comparing values}
best_rmse %>% 
  bind_rows(best_rmse_pct_loss, 
            best_rmse_one_std_err, 
            best_r2, 
            best_r2_pct_loss, 
            best_r2_one_std_error)
```

## 5. Final Specification

```{r final_spec_fit}
final_spec <- boost_tree(
  trees = best_r2$trees,           # Number of boosting rounds (trees)
  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
  min_n = best_r2$min_n,           # Minimum number of samples to split a node
  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```

## 6. Final Fit and Predictions

## Validation

```{r final_fit}
set.seed(10)
final_fit <- last_fit(final_spec,
                weather_recipe,
                split = weather_split)

final_fit %>%
  collect_predictions()
```

## 7. Evaluate on *Test Set*

```{r final_fit_metrics}
final_fit %>%
  collect_metrics()
```

## 8. Evaluate on Training Set

```{r}
final_spec %>%
  fit(yield ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rmse(yield, .pred) %>%
  bind_rows(

# R2
final_spec %>%
  fit(yield ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rsq(yield, .pred))
```

## 9. Predicted vs Observed Plot

```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous() +
  scale_y_continuous() 
```

## 10. Variable Importance

```{r final_spec}
final_spec %>%
  fit(yield ~ .,
         data = bake(weather_prep, weather_train)) %>% #There little change in variable improtance if you use full dataset
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)
```

# End of XGboost
